Untitled

### 2025-10-10 output/Response_to_action_editor_reviewers-2.Rmd

###### 3) I would like to see additional comments in their closing paragraph on how testing practitioners should use or what insights can be obtained by reviewing the different representations of the latent ability space and the item space. I see the mention of multigroup equating/linking, but I would like to see more discussion about how the test results and subsequent uses could be affected.

<!-- # TODO: Discuss complex items, and the proper representation in the test space. -->
<!-- # TODO: Discuss "latent space" transformation is an artifact, not something to "explore". -->

In addition to the previous changes, we have made a minor addition in the "Covariance-based version" section, to refer to the transformation that turns the latent trait distribution into a standard normal by its usual name in the literature, i.e., a "whitening transformation".
The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:

> Therefore, an orthonormal basis should represent independent,
> standard coordinates.
> Formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
> $`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$,
> which implies that Equation (2) actually represents
> a *whitening transformation* [@fukunaga_introduction_2013].
> Given this, $`r COV_MATRIX_TRANSF_EQ`$, and thus

@@

- The second paragraph in section "Graphical Representation" now clarifies better that the representation in the test space must use its basis instead of the canonical basis, to make the plot be a faithful representation of the actual directions of the items with respect to the latent trait dimensions.
This paragraph now reads (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):

> [...] Coordinates in a rectangular Cartesian system are commonly assumed
> to be simply represented in the canonical basis.
> To represent an item in the test space properly, however,
> we must use its basis $`r TEST_SPACE_BASIS`$,
> which implies pre-multiplying its coordinates
> by $`r TRANSFORM_MATRIX_TRANSP_INV`$
> to find the equivalent coordinates in the canonical basis. [...]

- The section "Graphical Representation Example" has been expanded to explain better both Table 1 and Figure 1.
Starting in the first paragraph (p. <!-- # TODO: Complete -->, l. <!-- # TODO: Complete -->), this section now reads:

> [...] The example in Figure 1 showcases a representation of
> a set of items with two different latent trait distributions.
> (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
> for simplicity, but note that they represent the coordinates of
> a latent trait point expressed in test space coordinates,
> that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
> Their parameters are shown in Table 1,
> both in the original `r MODEL_ACRONYM` metric and
> in the multidimensional metric under the two cases,
> which differ in the latent trait correlation.
> In the first case (columns $`r CORR_ORTH_OUT`$),
> the two latent dimensions are independent;
> in the second one ($`r CORR_OBL_OUT`$),
> the correlation between them is .5.
> The same `r MODEL_ACRONYM` parameters are used in both cases,
> but their (correlation-based) multidimensional parameters change
> with the correlation.
> If we used the agnostic version instead, the multidimensional parameters,
> given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.
> 
> Panel (a) of Figure 1 shows
> the item representation in the uncorrelated case,
> while the $`r CORR_OBL_OUT`$ case is shown in panels (b) to (d)
> under different conditions:
> with item coordinates computed from the agnostic version of the parameters
> (panel (b)), or from the correlation-based version,
> either using an appropriate oblique basis (panel (c)),
> or plotting them directly in the canonical basis (panel (d)).
> Keeping the horizontal axis invariant in panel (c)
> allows for a better comparison with the other plots;
> therefore, we have chosen a $`r TRANSFORM_MATRIX_TRANSP_INV`$ that
> only rotates the vertical axis [@harman_modern_1976]:
> 
> \begin{equation}
>   `r TRANSFORM_MATRIX_EXAMPLE_EQ`.
> \end{equation}
> 
> In all four panels, the contours of a bivariate standard normal density
> with the corresponding correlation are plotted along with the items
> (from outer to inner, the contours are the `r contour_breaks_out` of
> the density at its mode).
> The stretched ellipses in panel (c) are the result of transforming
> the distribution in the latent space to the test space,
> according to Equation (24).
> This implies that the covariance matrix of the distribution projected into
> the test space is transformed inversely as in the latent space, i.e.,
> $`r COV_MATRIX_TRANSF_TS_EQ`$.
> 
> Panels (a) and (b) show that,
> even if the agnostic coordinates are the same in both cases,
> their relative position with respect to the latent trait distribution is not
> ---this is the effect of disregarding the covariance structure.
> The correlation-based coordinates, plotted in panels (c) and (d),
> are also different in general from their agnostic counterpart (panel (b)).
> Although they do not vary in some cases (items 1 and 4),
> the difference is especially salient in the location of items 2 and 3,
> which are close to the 90% contour in the agnostic version
> but almost on the 50% contour in the correlation-based one.
> Panel (d) also illustrates the effect of plotting the coordinates directly
> on the canonical basis;
> the item locations and their positions relative to the distribution
> are correctly represented (which makes comparing with panel (b) easier),
> but their discrimination parameters and angles with the axes are distorted:
> For example, items 4 and 5 seem to be equally discriminating,
> when Table 1 panel (c) show they are not;
> and item 1 forms 45-degree angles with both axes,
> instead of the 30-degree ones shown in Table 1.
> 
> A comparison of panels (a) and (c) allows understanding the effect of
> a positive correlation on the $`r MDISC_CORR_PARAM`$
> (and consequently the $`r MIL_CORR_PARAM`$) value: [...]

### 2025-10-10 output/Response_to_action_editor_reviewers-2.Rmd

###### 2) I would like to see the graphics improved as suggested by Thissen's Review 2.  Labeling of axes is key for clarity.  And providing contour plots of the factor score ( latent variable) distributions will also help the reader’s understanding, especially for those readers coming from the IRT perspective.

We have made the changes proposed, updated the figure caption accordingly, and added an appropriate footer.
These changes have required a rather high effort to make sure they represent the algebra adequately, as well as some edits to improve the clarity of the plots and justify the computations.
Please see our detailed responses to Thissen's comments below.

###### 3) I would like to see additional comments in their closing paragraph on how testing practitioners should use or what insights can be obtained by reviewing the different representations of the latent ability space and the item space. I see the mention of multigroup equating/linking, but I would like to see more discussion about how the test results and subsequent uses could be affected.

We have tried our best to give a satisfactory response to this comment.
Although we acknowledge that we are not completely sure we have succeeded, we have carefully thought about the broader implications to our results, and made changes to our "Discussion" section to address this comment.
Besides some minor changes to adapt our discussion (please refer to the tracked changes in the main text of the submission), we have made two major changes.
The first one is the inclusion of additional comments to discuss the implications in terms of the test results and how their uses could be affected.
They have been included in what was previously the second paragraph;
we consider this text was misplaced in our previous version and have moved it accordingly, nearly to the end of the section.
After the changes, this paragraph (now in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) reads:

> We have seen that taking the latent space structure into account may have a
> substantial effect in the resulting values and
> interpretation of the parameters, something not contemplated by the original
> (hereby referred to as agnostic) version.
> Interpreting the item properties correctly requires a solid understanding of
> their relationship with the latent covariance structure,
> and we have shown that this can only be achieved
> with a proper geometrical representation.
> Simple items are not affected much by this issue but,
> as MIRT evolves and converges with the common factor psychometric tradition,
> we find MIRT models applied more often to new problems,
> involving complex multidimensional item structures
> (even including items with opposite-signed loadings).
> Our results will be especially relevant in those cases,
> where the interpretation of the items may be affected more drastically.
> For example, the relative importance of an item in scoring a given trait
> could be downplayed if it is strongly aligned with that trait dimension
> but this alignment is overlooked.
> Once again, in multi-group applications,
> where differences in the latent space structure is critical,
> the agnostic version may overlook relevant differences in
> the values of the multidimensional parameters.
> Different covariance structures across groups may significantly affect them
> due to differences in either the variances or the correlation structure.
> Therefore, even when the model parameters are equal across groups,
> the multidimensional parameters may not be the same
> if the covariance structures differ.
> Conversely, the same multidimensional parameters may correspond
> to different parameter models.
> The subsequent invariance of either the M2PL parameters
> or the multidimensional ones may lead to different interpretations of
> the same model when applied to different groups.
> Although this is an empirical question that requires further research,
> it could have significant implications for future theoretical developments
> in the understanding of MIRT theory.

Regarding the insights that can be obtained from the different representations,
we hope that the changes to Figure 1 and the subsequent additions
should account for how the test space representation should be interpreted.
Regarding the latent space, our latest version already stated that "the axes defined by the test space basis, and not the ones defined by the latent space basis, are a geometrical representation of the covariance structure. Indeed, the cosines of the angles among the test space axes are the geometric equivalent of the correlations. This result is both surprising and counterintuitive [...]".
The manuscript implicitly showed that the latent space transformation is as an "geometric artifact", necessary for arriving at the proper representation of the items in the test space.
We acknowledge though that this might not be necessarily clear in the previous submission.
The second change in the discussion consists of stating this explicitly now at the end of the fourth paragraph (previously the fifth one) of the "Discussion" section, in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->:

> [...] In fact, we have only defined the latent space as a geometrical artifact
> for obtaining the multidimensional parameters.
> Interpreting the items in the latent space
> could only lead researchers to wrong conclusions. [...]

Besides the aforementioned changes, we would like to highlight a minor addition to the "Covariance-based version" section, to refer to the transformation of the latent trait distribution into another one with an identity covariance matrix by its usual name in the literature, i.e., a "whitening transformation".
The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:

@@

> Therefore, an orthonormal basis should represent independent,
> standard coordinates.
> Formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
> $`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$,
> which implies that Equation (2) actually represents
> a *whitening transformation* [@fukunaga_introduction_2013].
> Given this, $`r COV_MATRIX_TRANSF_EQ`$, and thus

(Please note also the update in the title page to acknowledge the colleague who brought this previous literature to our attention, as well as the contributions of the reviewers.)

### 2025-10-22 notebooks/ESM_latent-space-representation.Rmd
Output the latent space plot in SVG format

ROOT_DIR <- here()

@@

```{r items-latent-space-plot-out}
#| cache:          false
#| fig.topcaption: true
#| fig.height:     4.7
#| fig.width:      6.65
#| warning:        false
plot_oblique_ls
```

>>>>

ROOT_DIR <- here()
DOC_DIR  <- file.path(ROOT_DIR, "output")

@@

```{r items-latent-space-plot-out}
#| cache:          false
#| fig.topcaption: true
#| fig.height:     4.7
#| fig.width:      6.65
#| warning:        false
plot_oblique_ls

ggsave(file.path(DOC_DIR, "Figure_ESM.svg"), width = 6.65, height = 4.7)
```
